---
title: "scRNAseq Interstitial cells : What happens over time?"
author: "Marion Hardy"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    theme: spacelab
    highlight: monochrome
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, cache = T, echo = FALSE, warning = F, cache.lazy = F)
knitr::opts_chunk$set(fig.width=6, fig.height=4) 

library(tidyverse)
library(openxlsx)
library(Seurat)
library(scCustomize)
```

# Introduction

Data from our collaborator Panagiotis rds file for a
SingleCellExperiment object containing the single cell data for **the
interstitial cells** of *Hydra Vulgaris* during multiples stages of
regeneration after bisection:

<https://www.dropbox.com/scl/fi/dg76sigjnj5u6qr06xk79/sce_interstitial_Juliano.rds?rlkey=f0y3lqt0wdcwq652zisnrpf9t&dl=0>

BUT they mapped it to *Hydra Magnipapillata (102)* "Drop-seq reads from
15 libraries generated for Hydra vulgaris strain AEP were mapped to the
2.0 genome assembly of closely related Hydra vulgaris strain 102
(available at <https://research.nhgri.nih.gov/hydra/>) and processed
using the Hydra 2.0 gene models. Strain Hydra vulgaris 102 was formerly
referred to as Hydra magnipapillata."

The coldata of the object contain cell annotation including

-   quality metrics: nFeature nCount (not MT percentage interestingly)

-   batch info: either 2869 (3162 barcodes), 3113 (10352 barcodes), 3271
    (13279 barcodes), 3357 (3875 barcodes)

-   originating experiments (head or foot regeneration)

-   experimental time points

-   pseudo-axis assignment (vals.axis ranging from 0-1, increasing in
    the foot-tentacle direction)

-   mitotic and apoptotic signatures indices from 0 to 1

The rowdata contains gene annotation, using Entrez-gene identifiers. I
have also noticed that in the sce objects there's

-   PCA, tSNE and UMAP coordinates for reduced dimensions + corrected
    for batch values

-   assay metafeatures hold gene_id, product, gene, is.rib.prot.gene
    (T/F), HypoMarkers (T/F), ccyle (T/F), apopt (T/F) etc

I converted the sce objects 6.6gb into a seurat object 2.2 gb and
checked that all cited parameters could be found in it.

# Summary of talks with Hannah and Celina

-   REG_FOOT = It's a head that is regenerating a foot (counterintuitive)
-   Clusters have been attributed with confidence except 10 which is likely doublets but should be explored further
-   Cluster 5 are probably glands but I haven't found a characterized marker I could use with the Hydra atlas
-   It looks like the feet neurons were not captured in this dataset :(
-   Explore their pseudoaxis parameter maybe?? see the preprint (https://www.biorxiv.org/content/10.1101/2024.02.08.579449v1.full.pdf)


# Data loading and upate on the project

In this report we're gonna plot transcription factors of interest over the different timepoints.
I also wanted to add cell population counts over time.
Maybe I could learn to plot gene expression over time too?=
Anyway, we're exploring gene expression programs and cell populations over time.

```{r}
# seurat objects
seurat_f = readRDS("./data_output/interstitial/seurat_filtered.rds")
# transcription factors of interest
tr = read.xlsx("./data/mcbi_dataset_MH_annotated.xlsx", sheet = "transcription")

```

```{r}
# changing labeled time from t6 to t06 so that it orders properly in the plots

seurat_f@meta.data =
  seurat_f@meta.data %>%
  mutate(Timepoint =
           ifelse(EXP_TIME == "REG_FOOT_t6", "REG_FOOT_t06",
                         ifelse(EXP_TIME == "REG_HEAD_t6", "REG_HEAD_t06", EXP_TIME)))

```

# Subsetting by head/foot and timepoint
## Not batch regressed

```{r}

Idents(seurat_f) = "Timepoint"
DimPlot(seurat_f, label = FALSE)

# Creating the head and foot data subset
hfootf = subset(seurat_f, subset = EXP == "REG_FOOT")
hheadf = subset(seurat_f, subset = EXP == "REG_HEAD")

```

```{r, fig.width=30, fig.height=6}

# Plotting conditions separately

Idents(hfootf) = hfootf@meta.data$SCT_snn_res.0.025
Idents(hheadf) = hheadf@meta.data$SCT_snn_res.0.025

DimPlot(hfootf, reduction = "umap", split.by = "Timepoint", pt.size = 1)

DimPlot(hheadf, reduction = "umap", split.by = "Timepoint", pt.size = 1)

```

# Cluster attribution

```{r, fig.width=8, fig.height=6}

Idents(seurat_f) = "attr_clusters"

DimPlot(seurat_f, label = TRUE)+
  labs(title = "Attributed clusters")

```

# Transcription marker expression

Celina provided me with a list of markers of interest that I tried to find in the 102 genome.

```{r}

tr %>% 
  select(Symbol, Symbol_updated)

```

```{r}

uh = DotPlot_scCustom(seurat_f, features = tr$Symbol_updated, 
                      colors_use = viridis_plasma_dark_high,
                      x_lab_rotate = T)+
  labs(title = "Genes from the homeostatic animal")

uh
```

```{r, fig.height=10, fig.width=15}

FeaturePlot(object = seurat_f, 
            features = tr$Symbol_updated,
            reduction = "umap",
            order = TRUE,
            min.cutoff = 'q10', 
            label = F,
            repel = TRUE,
            pt.size = 0.75)

```

# Pseudo-axis

```{r}

Idents(seurat_f) = "vals.axis_k25"

DimPlot(seurat_f, label = TRUE)+
  labs(title = "Pseudo-axis values")

```

# Scaling down to only neurons 

I am scaling down to see if I can get genes with interesting patterns between stem cells -> neurons
I am trying to see regulatory genes appear with the appearing head neuron cluster...

```{r, eval=FALSE}

neurons = subset(seurat_f, subset = attr_clusters %in% c("ISC/earlyNem","ec1A/B",
                                                         "ec2","ec4","Ec3N/En1N"))  
Idents(neurons) = "attr_clusters"

DimPlot(neurons, label = TRUE)+
  labs(title = "Pseudo-axis values")

```

## Reprocessing for re-clustering

```{r}

neurons <- RunPCA(neurons, features = VariableFeatures(object = neurons))
DimPlot(neurons, reduction = "pca")

neurons <- RunUMAP(neurons, dims = 1:28, seed.use = 054057)
DimPlot(neurons, reduction = "umap", label = T)

neurons <- FindNeighbors(neurons, dims = 1:28)

neurons <- FindClusters(neurons, resolution = 0.025)
DimPlot(neurons, label = TRUE)+
  labs(title = "Resolution = 0.025")

```

```{r}
# Creating the head and foot data subset
hfootf = subset(neurons, subset = EXP == "REG_FOOT")
hheadf = subset(neurons, subset = EXP == "REG_HEAD")

```

```{r, fig.width=30, fig.height=6}

# Plotting conditions separately

Idents(hfootf) = hfootf@meta.data$SCT_snn_res.0.025
Idents(hheadf) = hheadf@meta.data$SCT_snn_res.0.025

DimPlot(hfootf, reduction = "umap", split.by = "Timepoint", pt.size = 1)

DimPlot(hheadf, reduction = "umap", split.by = "Timepoint", pt.size = 1)

```


```{r, fig.height=10, fig.width=15}

FeaturePlot(object = neurons, 
            features = tr$Symbol_updated,
            reduction = "umap",
            order = TRUE,
            min.cutoff = 'q10', 
            label = F,
            repel = TRUE,
            pt.size = 0.75)

```



```{r, eval = FALSE}
write_rds(seurat_f, "./data_output/interstitial/seurat_filtered.rds")
```

# Session info

```{r}
sessionInfo()
```
